Rapid-response Bayesian sky localization for electromagnetic follow-up of gravitational-wave candidates

### 2013-06-20 ms.tex
Fine-tune some wording in introduction
Actually, I don't know what order this integration scheme is
Fixed typo in equation
Add full-size parens
Finished describing adaptive integration scheme

In this paper we present a rapid and accurate method of sky localization that makes use of Bayesian methods. It differs from existing techniques is several important ways.  In the first place, we fix the intrinsic parameters to their \ac{ML} estimates, as provided by the detection pipeline.  This reduces the dimensionality of the parameter we need to search. In addition, the technique takes as its input the \ac{ML} estimates of the extrinsic parameters, instead of the $h(t)$ time series. This makes the likelihood itself much faster to evaluate. Finally, instead of of using \ac{MCMC} or some similar method for statistical sampling, we make use of a deterministic quadrature scheme. We call this technique \ac{BAYESTAR}%
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \acp{GW} detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `T' in the acronym, because then it would be called BAYES\nobreakdashes-TART, which would sound ridiculous.}%
. It is unique in that in bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. We expect that \ac{BAYESTAR} will take on a key role in observing \ac{CBC} events in both \ac{GW} and optical channels during the Advanced \ac{LIGO} era.

@@

The marginalization over $D_\mathrm{L}$, $u$, and $\psi$ must be done numerically. The outermost integration is over $u$ and $\psi$, and is done with Newton\nobreakdashes--Cotes by evaluating $\mathcal{L}_\mathrm{SNR}$ on a regular lattice in $u$ and $\psi$. This integration scheme is first order in $u$ and second order in $\psi$.

\marginpar{Explain initial integration breakpoints and how they are calculated. Cite \texttt{gsl\_integrate\_qagp}.}%
%
The innermost integration is over $D_\mathrm{L}$ and is performed using adaptive Gaussian quadrature, but with a twist that speeds up convergence.

Before evaluating each inner integral over $D_\mathrm{L}$, the quantities $A=-\frac{1}{2} \sum_i {\rho_i}^2$ and $B=\sum_i \rho_i \hat\rho_i$ are precalculated so that the integrand can be written as
%
\begin{equation*}
	\exp(A {D_\mathrm{L}}^{-2} + B D_\mathrm{L}) {D_\mathrm{L}}^m.
\end{equation*}
%
Then, we see that the likelihood $\exp(A {D_\mathrm{L}}^{-2} + B D_\mathrm{L})$ is maximized when $D_\mathrm{L} = D_{\mathrm{L},0} = -2A/B$. The likelihood takes on a factor $\eta$ of its maximum value when
%
\begin{equation}
	D_\mathrm{L} = D_{\mathrm{L},\pm} = \left(\frac{1}{D_{\mathrm{L},0}} \pm \sqrt{\frac{\log\eta}{A}}\right)^{-1}.
\end{equation}

\begin{figure}
    \caption{\label{fig:radial_integrand}Illustration of initial subdivisions for radial integration scheme.}
    \begin{center}
        \includegraphics{radial_integrand}
    \end{center}
\end{figure}

>>>>

In this paper we present a rapid and accurate method of sky localization that makes use of Bayesian methods. It differs from existing techniques is several important ways.  In the first place, we fix the intrinsic parameters to their \ac{ML} estimates, as provided by the detection pipeline.  This reduces the dimensionality of the parameter space we need to sample. In addition, the technique takes as its input the \ac{ML} estimates of the extrinsic parameters, instead of the $h(t)$ time series. This makes the likelihood itself much faster to evaluate. Finally, instead of of using \ac{MCMC} or some similar method for statistical sampling, we make use of a deterministic quadrature scheme. We call this technique \ac{BAYESTAR}%
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \acp{GW} detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `T' in the acronym, because then it would be called BAYES\nobreakdashes-TART, which would sound ridiculous.}%
. It is unique in that in bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. We expect that \ac{BAYESTAR} will take on a key role in observing \ac{CBC} events in both \ac{GW} and optical channels during the Advanced \ac{LIGO} era.

@@

The marginalization over $D_\mathrm{L}$, $u$, and $\psi$ must be done numerically. The outermost integration is over $u$ and $\psi$, and is done with Newton\nobreakdashes--Cotes by evaluating $\mathcal{L}_\mathrm{SNR}$ on a regular lattice in $u$ and $\psi$.

\begin{figure}
    \caption{\label{fig:radial_integrand}Illustration of initial subdivisions for distance integration scheme. Distance increases from left to right. In the print version, the left\nobreakdashes-hand tail, the left\nobreakdashes- and right\nobreakdashes-hand sides of the maximum likelihood peak, and the right\nobreakdashes-hand tail, are colored cyan, red, green, and blue, respectively.}
    \begin{center}
        \includegraphics{radial_integrand}
    \end{center}
\end{figure}

The innermost integration is over $D_\mathrm{L}$ and is performed using adaptive Gaussian quadrature, but with a twist that speeds up convergence. As in many parameter estimation problems, the integrand may be sharply peaked, and any generic adaptive integration scheme may take a long time to find the peak. Before evaluating each inner integral over $D_\mathrm{L}$, the quantities $A=-\frac{1}{2} \sum_i {\rho_i}^2$ and $B=\sum_i \rho_i \hat\rho_i$ are precalculated so that the integrand can be written as
%
\begin{equation*}
	\exp\left(\frac{A}{{D_\mathrm{L}}^2} + \frac{B}{D_\mathrm{L}}\right) {D_\mathrm{L}}^m.
\end{equation*}
%
Then, we see that the likelihood $\exp(A {D_\mathrm{L}}^{-2} + B {D_\mathrm{L}}^{-1})$ is maximized when $D_\mathrm{L} = D_{\mathrm{L},0} = -2A/B$. The likelihood takes on a factor $\eta$ (say, $\eta=0.01$) of its maximum value when
%
\begin{equation}
	D_\mathrm{L} = D_{\mathrm{L},\pm} = \left(\frac{1}{D_{\mathrm{L},0}} \mp \sqrt{\frac{\log\eta}{A}}\right)^{-1}.
\end{equation}
%
We have now identified up to five breakpoints that partition the distance integrand into up to four intervals with quantitatively distinct behavior. These intervals are depicted in Figure~\ref{fig:radial_integrand} with distance increasing from left to right. There is a left\nobreakdashes-hand or small distance tail in which the integrand is small and monotically increasing, a left\nobreakdashes- and right\nobreakdashes-hand side of the maximum likelihood peak, and a right\nobreakdashes-hand tail in which the integrand is small and monotonically decreasing. These breakpoints are:
%
\begin{equation}
    D_{\mathrm{L},\mathrm{break}} = \{ D_\mathrm{L} \in
    \left\{
    \begin{array}{c}
    D_{\mathrm{L},\mathrm{min}} \\
    D_{\mathrm{L},-} \\
    D_{\mathrm{L},0} \\
    D_{\mathrm{L},+} \\
    D_{\mathrm{L},\mathrm{max}}
    \end{array}
    \right\} :
    D_{\mathrm{L},\mathrm{min}} \leq D_\mathrm{L}
    \leq D_{\mathrm{L},\mathrm{max}}\}.
\end{equation}
%
We use these breakpoints as initial subdivision in an adaptive Gaussian quadrature algorithm. Specifically, we use the GNU Scientific Library's \verb|gsl_integrate_qagp|\footnote{\url{http://www.gnu.org/software/gsl/manual/html_node/QAGP-adaptive-integration-with-known-singular-points.html}} function, which is designed to integrate functions with known singular points (though, of course, or integrand has no singular points). This function estimates the integral over each subdivision and each interval's contribution to the total error, then subdivides the interval whose error contribution is largest. Subdivisions continue until a fixed total fractional error is reached. In this way, most integrand evaluations are expended on the most important distance interval, whether that happens to be the tails (when the posterior is dominated by the prior) or the peak (when the posterior is dominated by the observations).
